---
title: "Assembling and summarizing school trips"
output:
  html_document:
    theme: simplex
    toc: yes
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The purpose of this notebook is to assemble a sample of trips to school from the
2009 and 2017 samples of National Household Travel Survey (NHTS) in the United 
States and present some basic descriptive statistics of that combined sample.

# Required packages

The analysis uses the following packages:

```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)
library(knitr)
library(janitor)
```

# Download data

At this time (March 7, 2025), public versions of NHTS data (without respondents'
personally-identifiable information) are available to download from 
https://nhts.ornl.gov/downloads.


```{r, eval=FALSE}
# 2022 data
temp <- tempfile()
download.file("https://nhts.ornl.gov/assets/2022/download/csv.zip",temp)
people22 <- read_csv(unz(temp, "perv2pub.csv"))
trip22 <- read_csv(unz(temp, "tripv2pub.csv")) 
unlink(temp)

# 2017 data
temp <- tempfile()
download.file("https://nhts.ornl.gov/assets/2016/download/csv.zip",temp)
people17 <- read_csv(unz(temp, "perpub.csv"))
trip17 <- read_csv(unz(temp, "trippub.csv"))
unlink(temp)

# 2009 data
temp <- tempfile()
download.file("https://nhts.ornl.gov/assets/2009/download/Ascii.zip",temp)
people09 <- read_csv(unz(temp, file.path("Ascii","PERV2PUB.CSV")))
trip09 <- read_csv(unz(temp, file.path("Ascii","DAYV2PUB.CSV"))) 
unlink(temp)

```

Or, if you already have the data downloaded, you can load it from file.

```{r, message=FALSE}

# 2022 data
people22 <- here("data",
                 "nhts",
                 "2022",
                 "perv2pub.csv") |>
  read_csv()

trip22 <- here("data",
               "nhts",
               "2022",
               "tripv2pub.csv") |>
  read_csv() 

# 2017 data
people17 <- here("data",
                 "nhts",
                 "2017",
                 "perpub.csv") |>
  read_csv()

trip17 <- here("data",
               "nhts",
               "2017",
               "trippub.csv") |>
  read_csv() 

# 2009 data
people09 <- here("data",
                 "nhts",
                 "2009",
                 "PERV2PUB.CSV") |>
  read_csv()

trip09 <- here("data",
               "nhts",
               "2009",
               "DAYV2PUB.CSV") |>
  read_csv() 
```


# Combine samples

We'll combine the data across the three years.

## People

```{r}
people22 <- people22 |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2022) |>
  rename(sex = R_SEX_IMP) 

people17 <- people17 |> 
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX_IMP,
         WORKER,
         HHSIZE) |>
  mutate(year = 2017) |>
  rename(sex = R_SEX_IMP)

people09 <- people09 |>
  select(HOUSEID, 
         PERSONID,
         R_AGE,
         R_SEX,
         WORKER,
         HHSIZE) |>
  mutate(year = 2009) |>
  rename(sex = R_SEX)

all_people <- rbind(people09, people17) |>
  rbind(people22) |>
  mutate(person_hh_year = paste(PERSONID, HOUSEID, year, sep = "-")) |>
  mutate(hh_year = paste(HOUSEID, year, sep = "-"))
```


## Trips

The 2022 dataset does not include a density variable, so we're creating one just
and tagging it with a value of 2022 for now. We will filter the 2022 data back 
out before we finalize the dataset for the discrete choice model.

```{r}
trip09 <- trip09 |>
  select(-ONTD_P15, -ONTD_P14) |>
  mutate(year = 2009,
         mode = case_when(TRPTRANS == "22" ~ "Bike",
                          TRPTRANS == "23" ~ "Walk", 
                          TRPTRANS %in% c("01", "02", "03", "04", "05")~ "Car",
                          TRPTRANS == "11" ~ "School bus",
                          TRPTRANS %in% c("09", "10", "12", "13", "14", 
                                          "15", "16", "17", "18") ~ "Transit",
                          TRUE ~ "Other"),
         home_to_school = (WHYFROM == "01" & WHYTO == "21")) |>
  rename(sex = R_SEX,
         density = HBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         home_to_school,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         TRPTRANS,
         HHC_MSA)

trip17 <- trip17 |>
  mutate(year = 2017,
         mode = case_when(TRPTRANS == "01" ~ "Walk",
                          TRPTRANS == "02" ~ "Bike",
                          TRPTRANS %in% c("03", "04", "05", "06") ~ "Car",
                          TRPTRANS == "10" ~ "School bus",
                          TRPTRANS %in% c("11", "12", "13", "14", "15", "16") ~ "Transit",
                          TRUE ~ "Other"),
         home_to_school = (WHYFROM %in% c("01","02") & WHYTO == "08"),
         ONTD_P10 = as.character(ONTD_P10))|>
  mutate(MSA = "unknown") |>
  rename(sex = R_SEX_IMP,
         density = OBPPOPDN) |>
  select(PERSONID,
         HOUSEID,
         home_to_school,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         TRPTRANS) 

trip22 <- trip22 |>
  mutate(year = 2022,
         mode = case_when(TRPTRANS == "20" ~ "Walk",
                          TRPTRANS == "18" ~ "Bike",
                          TRPTRANS %in% c("01", "02", "03", "04", "06") ~ "Car",
                          TRPTRANS == "09" ~ "School bus",
                          TRPTRANS %in% c("10", "11", "12", "13", "17") ~ "Transit",
                          TRUE ~ "Other"),
         home_to_school = (WHYFROM %in% c("01","02") & WHYTO == "06"),
         density = 2022,
         ONTD_P10 = as.character(ONTD_P10)) |>
  mutate(MSA = "unknown") |>
  rename(sex = R_SEX_IMP) |>
  select(PERSONID,
         HOUSEID,
         home_to_school,
         year,
         R_AGE,
         ENDTIME,
         TRPMILES,
         mode,
         starts_with("ONTD_P"),
         NUMONTRP,
         NONHHCNT,
         sex,
         HHFAMINC,
         DRVRCNT,
         HHVEHCNT,
         density,
         TRPTRANS)
         

all_trips <- bind_rows(trip09, trip17, trip22)

```


# Summarize sample sizes by filter and year

The criteria for including a trip in the model are:

* Traveler is between the ages of 7 and 14
* Trip qualifies as a trip to school
    * Trip begins at home and ends at school
    * Traveler arrives at school before 10am
    * Trip is the first such trip of the day
    * Trip distance is greater than zero
* No missing data for predictor variables
    * Density is not missing
    * Household income is not missing
* Trip was a reasonable walking distance (less than 1.25 miles)
* Trip was by a mode that was likely to be unavailable to most students
    * Trip was not by school bus
    * Trip was not by public transit
    * Trip was not by other modes
    
## Total sample size

We'll start by counting the number of households, people, and trips in each 
survey year.

```{r, message=FALSE}
all_trips <- all_trips |>
  mutate(is_child = R_AGE > 6 & R_AGE < 15,
         HOUSEID = as.character(HOUSEID)) 

all_people <- all_people |>
  mutate(is_child = R_AGE > 6 & R_AGE < 15)

all_households <- all_people |>
  group_by(HOUSEID, year) |>
  summarise(has_child = sum(is_child) > 0) |>
  ungroup()

hh_summary <- all_households |>
  group_by(year) |>
  summarise(n = n()) |>
  mutate(`Unit of observation` = "Households")

people_summary <- all_people |>
  group_by(year) |>
  summarise(n = n()) |>
  mutate(`Unit of observation` = "People")

trips_summary <- all_trips |>
  group_by(year) |>
  summarise(n = n()) |>
  mutate(`Unit of observation` = "Trips")

full_sample <- bind_rows(hh_summary,
                         people_summary,
                         trips_summary) |>
  mutate(Sample = "Full survey sample") |>
  group_by(year) 

```

## Sample size of children

Then we'll count the number of children in our target age range by survey year.

```{r, message=FALSE}
child_trips <- all_trips |>
  filter(is_child) 

child_people <- all_people |>
  filter(is_child)

child_hh <- all_households |>
  filter(has_child)

child_hh_summary <- child_hh |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Households")

child_people_summary <- child_people |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "People")

child_trips_summary <- child_trips |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Trips")

child_sample <- bind_rows(child_hh_summary,
                          child_people_summary,
                          child_trips_summary) |>
  mutate(Sample = "Limit to children ages 7 to 14") |>
  group_by(year) |>
  left_join(full_sample[c("year", "Unit of observation", "n")]) |>
  mutate(reduction = 1-n_new/n) |>
  select(year, n_new, `Unit of observation`, Sample, reduction) |>
    rename(n = n_new) 

```

## Sample size of children who attended school

Of those, how many had a school trip on the survey day?

Again, a school trip is one that:

* begins at home and ends at school,
* arrives at its destination before 10am, and
* is the first such trip of the day.

```{r, message=FALSE}

school_trips <- child_trips |>
  mutate(arrival = as.numeric(ENDTIME)) |>
  filter(home_to_school, 
         arrival < 1000) |>
  group_by(PERSONID, HOUSEID) |>
  summarise_all(first)

school_hhs <- school_trips |>
  group_by(HOUSEID, year) |> 
  summarize(n = n())

school_people <- school_trips |>
  group_by(HOUSEID, PERSONID, year) |> 
  summarize(n = n())

school_hh_summary <- school_hhs |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Households")

school_person_summary <- school_people |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "People")

school_trip_summary <- school_trips |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Trips")

school_sample <- bind_rows(school_hh_summary,
                          school_person_summary,
                          school_trip_summary) |>
  mutate(Sample = "Limit to school trips") |>
  group_by(year) |>
  left_join(child_sample[c("year", "Unit of observation", "n")]) |>
  mutate(reduction = 1-n_new/n) |>
  select(year, n_new, `Unit of observation`, Sample, reduction) |>
    rename(n = n_new) 

```

Our next filter will narrow this sample to only those trips within a reasonable
walking distance, but first, let's see what share of all trips are by school 
bus.

```{r, message=FALSE}
mode_summary <- school_trips |>
  filter(TRPMILES > 0) |>
  group_by(mode) |>
  summarize(n = n()) |>
  ungroup() |>
  mutate(`Percent of all school trips` = round(100*n/sum(n)))|>
  select(mode, `Percent of all school trips`) |>
  mutate(`Percent of all school trips` = 
           paste0(`Percent of all school trips`, "%") )
  
```


And let's break that down by whether the trip is longer or shorter than
1.25 miles.

```{r, message=FALSE}
mode_short_summary <- school_trips |>
  filter(TRPMILES > 0) |>
  mutate(short = ifelse(TRPMILES < 1.25, "short", "long")) |>
  mutate(short_year = paste0(short, year)) |>
  group_by(short_year, mode) |>
  summarize(n = n()) |>
  ungroup() |>
  group_by(short_year) |>
  mutate(pct = round(100*n/sum(n))) |>
  select(short_year, mode, pct) |>
  pivot_wider(names_from = short_year, values_from = pct) |>
  ungroup() |>
  arrange(desc(`long2017`)) |>
  mutate(`Percent of school trips shorter than 1.25 miles (2009)` = 
           paste0(`short2009`, "%"),
         `Percent of school trips shorter than 1.25 miles (2017)` = 
           paste0(`short2017`, "%"),
         `Percent of school trips shorter than 1.25 miles (2022)` = 
           paste0(`short2022`, "%"),
         `Percent of school trips longer than 1.25 miles (2009)` = 
           paste0(`long2009`, "%"),
         `Percent of school trips longer than 1.25 miles (2017)` = 
           paste0(`long2017`, "%"),
         `Percent of school trips longer than 1.25 miles (2022)` = 
           paste0(`long2022`, "%")) |>
  select(mode,
         `Percent of school trips shorter than 1.25 miles (2009)`,
         `Percent of school trips shorter than 1.25 miles (2017)`,
         `Percent of school trips shorter than 1.25 miles (2022)`,
         `Percent of school trips longer than 1.25 miles (2009)`,
         `Percent of school trips longer than 1.25 miles (2017)`,
         `Percent of school trips longer than 1.25 miles (2022)`) |>
  left_join(mode_summary) 

kable(mode_short_summary)
```


## Sample size of trips shorter than 1.25 miles

Of those school trips, how many were within a reasonable walking distance?

```{r, message=FALSE}
short_trips <- school_trips |>
  filter(TRPMILES < 1.25,
         TRPMILES > 0) 

short_hhs <- short_trips |>
  group_by(HOUSEID, year) |> 
  summarize(n = n())

short_hh_summary <- short_hhs |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Households")

short_trip_summary <- short_trips |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Trips")

short_sample <- bind_rows(short_hh_summary,
                          short_trip_summary) |>
  mutate(Sample = "Limit to trips shorter than 1.25 miles") |>
  group_by(year) |>
  left_join(school_sample[c("year", "Unit of observation", "n")]) |>
  mutate(reduction = 1-n_new/n) |>
  select(year, n_new, `Unit of observation`, Sample, reduction) |>
    rename(n = n_new) 

```


## Excluding school bus trips

School bus service is not available to all students, so we will exclude trips 
by students for whom it is an option.

```{r, message=FALSE}
no_bus_trips <- short_trips |>
  filter(mode != "School bus")

no_bus_hhs <- no_bus_trips |>
  group_by(HOUSEID, year) |> 
  summarize(n = n())

no_bus_hh_summary <- no_bus_hhs |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Households")

no_bus_trip_summary <- no_bus_trips |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Trips")

no_bus_sample <- bind_rows(no_bus_hh_summary,
                           no_bus_trip_summary) |>
  mutate(Sample = "Exclude school bus trips") |>
  group_by(year) |>
  left_join(short_sample[c("year", "Unit of observation", "n")]) |>
  mutate(reduction = 1-n_new/n) |>
  select(year, n_new, `Unit of observation`, Sample, reduction) |>
    rename(n = n_new)
```

## Excluding transit and other modes

And we will also exclude those who used transit or other less common modes.

```{r, message=FALSE}
no_other_mode_trips <- no_bus_trips |>
  filter(mode != "Transit",
         mode != "Other")

no_other_mode_hhs <- no_other_mode_trips |>
  group_by(HOUSEID, year) |> 
  summarize(n = n())

no_other_mode_hh_summary <- no_other_mode_hhs |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Households")

no_other_mode_trip_summary <- no_other_mode_trips |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Trips")

no_other_mode_sample <- bind_rows(no_other_mode_hh_summary,
                                  no_other_mode_trip_summary) |>
  mutate(Sample = "Exclude trips by uncommon modes") |>
  group_by(year) |>
  left_join(no_bus_sample[c("year", "Unit of observation", "n")]) |>
  mutate(reduction = 1-n_new/n) |>
  select(year, n_new, `Unit of observation`, Sample, reduction) |>
    rename(n = n_new)
```



## Sample size without missing data

Of those walking-distance school trips, how many had data available for all 
predictors?

```{r, message=FALSE}
non_missing_trips <- no_other_mode_trips |>
  filter(density != 2022,
         density > 0,
         HHFAMINC > 0) 

non_missing_hhs <- non_missing_trips |>
  group_by(HOUSEID, year) |> 
  summarize(n = n())

non_missing_hh_summary <- non_missing_hhs |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Households")

non_missing_trip_summary <- non_missing_trips |>
  group_by(year) |>
  summarise(n_new = n()) |>
  mutate(`Unit of observation` = "Trips")

non_missing_sample <- bind_rows(non_missing_hh_summary,
                                non_missing_trip_summary) |>
  mutate(Sample = "Exclude observations with missing data") |>
  group_by(year) |>
  left_join(no_other_mode_sample[c("year", "Unit of observation", "n")]) |>
  mutate(reduction = 1-n_new/n) |>
  select(year, n_new, `Unit of observation`, Sample, reduction) |>
    rename(n = n_new)

sample_trips <- non_missing_trips
```

## Full summary tables

Now we can combine all that data into a single table.

```{r}
full_summary <- bind_rows(full_sample,
                          child_sample,
                          school_sample,
                          short_sample,
                          no_bus_sample,
                          no_other_mode_sample,
                          non_missing_sample) |>
  pivot_wider(names_from = year, values_from = c("n", "reduction")) |>
  replace_na(list(n_2022 = 0)) |>
  mutate(Total = n_2009 + n_2017 + n_2022) 
```

First, let's see how the sample size changed from year to year with each
successive filter.

```{r, caption="Sample sizes by year and exclusion criteria"}
full_summary |>
  select(Sample,
         `Unit of observation`,
         n_2009,
         n_2017,
         n_2022,
         Total) |>
  rename(`2009` = n_2009,
         `2017` = n_2017,
         `2022` = n_2022) |>
  kable(format.args = list(big.mark = " "))


```

And then let's see the percentage by which each successive filter reduced the 
sample size.

```{r, caption="Sample sizes by year and exclusion criteria"}

full_summary |>
  select(Sample,
         `Unit of observation`,
         reduction_2009,
         reduction_2017,
         reduction_2022) |>
  rename(`2009` = reduction_2009,
         `2017` = reduction_2017,
         `2022` = reduction_2022) |>
  filter(Sample != "Full survey sample") |>
  replace_na(list(`2022` = 1)) |>
  kable(format.args = list(format = "f", digits = 1))
```


# Assemble data on child's household members.

We need information on the ages of the household members on each trip to 
determine whether a child was accompanied by a household adult.

We also need to determine whether each household includes a non-working female
adult and/or a non-working male adult and the ages of the youngest and 
oldest children in each household (to determine if the trip-maker has a 
younger or older sibling).

We'll reshape the person-level data into a household-level dataset with
a column indicating the age of each person in the household.

```{r}
hh_ages <- all_people |>
  select(hh_year, PERSONID, R_AGE) |>
  pivot_wider(names_from = PERSONID,
              names_prefix = "age_",
              values_from = R_AGE)
```

And here, we'll construct a variable for whether there is a no-working parent of 
either of two genders, and also for the ages of the oldest and youngest child
in each household.

```{r, message=FALSE}
who_in_hh <- all_people |>
  mutate(adult = R_AGE > 17,
         kid_age = ifelse(R_AGE < 18, R_AGE, -2),
         R_AGE = ifelse(R_AGE < 0, 999, R_AGE),
         non_worker_mom = R_AGE > 17 & WORKER != "01" & sex == "02",
         non_worker_dad = R_AGE > 17 & WORKER != "01" & sex != "02") |>
  group_by(hh_year, HHSIZE) |>
  summarize(num_records = n(),
         n_adults = sum(adult),
         non_work_mom = sum(non_worker_mom) > 0,
         non_work_dad = sum(non_worker_dad) > 0,
         min_age = min(R_AGE),
         max_kid_age = max(kid_age)) |>
  mutate(youngest_kid = ifelse(num_records < HHSIZE, 1, min_age),
         oldest_kid = ifelse(max_kid_age > 0, max_kid_age, 1),
         n_children = HHSIZE - n_adults) |>
  ungroup() |>
  filter(n_children > 0) |>
  select(hh_year, 
         non_work_dad,
         non_work_mom,
         oldest_kid, 
         youngest_kid)
```
And now we'll join that data to our sample of trips data.


```{r, message=FALSE}
sample_trips <- sample_trips |>
  mutate(hh_year = paste(HOUSEID, year, sep = "-")) |>
  left_join(hh_ages) |>
  left_join(who_in_hh)

```

# Construct variables

We want to include the following variables:

* `log_inc_k`: Income (log-transformed, continuous)
* `veh_per_driver` :Vehicles per driver
* `non_work_mom`: Presence of non-working adult woman in household
* `non_work_dad`: Presence of non-working adult man in household
* `age`: Age
* `female`: Sex
* `has_lil_sib`: Presence of younger child in household
* `has_big_sib`: Presence of older child in household
* `log_distance`: Distance
* `log_density`: Population density of home block group

And the outcome variable, `mode_ind`, which can take any of the following values:
    
* 18: Active mode accompanied by a household adult
* 28: Active mode unaccompanied by a household adult
* 17: In a car accompanied by a household adult
* 27: In a car unaccompanied by a household adult

We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1 (TRUE) for all respondents.

* av_kid_car
* av_par_car
* av_kid_act
* av_par_act

## Predictor variables

Our model(s) will include the following predictors:

* Household characteristics
  * Household income
  * Number of vehicles per driver
  * Presence of a non-working mother
  * Presence of a non-working father
* Individual characteristics
  * Age
  * Sex
  * Has a younger sibling
  * Has an older sibling
* Trip characteristics
  * Trip distance
  * Population density

### Construct income variable

The 2009 survey recorded household income in one of 
seventeen income categories and the 2017 survey recorded it in one of eleven 
income categories. We converted this to a continuous variable by assigning 
each household an income at the midpoint of its respective category. 
2009 incomes were increased by 14 percent to adjust for
inflation (converting them to 2017 dollars).

For the 
highest income category (greater than \$100,000 USD per year in the 2009 survey 
and greater than \$200,000 USD per year in the 2017), we calculated the median 
income of households in the United States with children between the ages of 
7 and 14 and incomes above the threshold, based on microdata from the American 
Community Survey published by IPUMS. The value assigned to the highest income 
category based on this method was \$140,000 for the 2009 survey and \$281,360 
for the 2017 survey. See https://github.com/vibe-lab-gsd/never-walk-alone/blob/main/code/upper-income-bin.R
for the code used to calculate those values.


```{r}
sample_trips <- sample_trips |>
  mutate(income_k = case_when(HHFAMINC == "01" & year == 2017 ~ 5,
                              HHFAMINC == "02" & year == 2017 ~ 12.5,
                              HHFAMINC == "03" & year == 2017 ~ 20,
                              HHFAMINC == "04" & year == 2017 ~ 30,
                              HHFAMINC == "05" & year == 2017 ~ 42.5,
                              HHFAMINC == "06" & year == 2017 ~ 62.5,
                              HHFAMINC == "07" & year == 2017 ~ 87.5,
                              HHFAMINC == "08" & year == 2017 ~ 112.5,
                              HHFAMINC == "09" & year == 2017 ~ 137.5,
                              HHFAMINC == "10" & year == 2017 ~ 175,
                              HHFAMINC == "11" & year == 2017 ~ 281.36,
                              HHFAMINC == "01" ~ 2.5*1.14,
                              HHFAMINC == "02" ~ 7.5*1.14,
                              HHFAMINC == "03" ~ 12.5*1.14,
                              HHFAMINC == "04" ~ 17.5*1.14,
                              HHFAMINC == "05" ~ 22.5*1.14,
                              HHFAMINC == "06" ~ 27.5*1.14,
                              HHFAMINC == "07" ~ 32.5*1.14,
                              HHFAMINC == "08" ~ 37.5*1.14,
                              HHFAMINC == "09" ~ 42.5*1.14,
                              HHFAMINC == "10" ~ 47.5*1.14,
                              HHFAMINC == "11" ~ 52.5*1.14,
                              HHFAMINC == "12" ~ 57.5*1.14,
                              HHFAMINC == "13" ~ 62.5*1.14,
                              HHFAMINC == "14" ~ 67.5*1.14,
                              HHFAMINC == "15" ~ 72.5*1.14,
                              HHFAMINC == "16" ~ 77.5*1.14,
                              HHFAMINC == "17" ~ 90*1.14,
                              HHFAMINC == "18" ~ 140*1.14,
                              TRUE ~ -9)) |>
  mutate(log_inc_k = log(income_k))
```


### Construct vehicles per driver variable

Calculate the number of vehicles per driver. If there are no drivers, use a 
value of zero.

```{r}
sample_trips <- sample_trips |>
  mutate(veh_per_driver = ifelse(DRVRCNT == 0, 0, HHVEHCNT / DRVRCNT))
```

### Rename age variable

Because why not?

```{r}
sample_trips <- sample_trips |>
  rename(age = R_AGE)
```

### Construct female variable

We're using a binary sex variable, where a value of one indicates that the
respondent either identified herself or was identified by the 
interviewer (imputed) as female.

```{r}
sample_trips <- sample_trips |>
  mutate(female = ifelse(sex == "02", 1, 0))
```

### Construct non-working parent variables

We're just converting the logical variable to a numeric variable so it works
in Biogeme.

```{r}
sample_trips <- sample_trips |>
  mutate(non_work_dad = ifelse(non_work_dad, 1, 0),
         non_work_mom = ifelse(non_work_mom, 1, 0))
```

### Construct sibling variables

Compare the child's age to the age of the oldest and youngest householdhold
child to determine if they have an older or younger sibling.

```{r}
sample_trips <- sample_trips |>
  mutate(has_lil_sib = ifelse(age != youngest_kid, 1, 0),
         has_big_sib = ifelse(age != oldest_kid, 1, 0))
```

### Construct distance and density variables

Log-transform distance and density.

```{r}
sample_trips <- sample_trips |>
  mutate(log_distance = log(TRPMILES),
         log_density = log(density))
```

### Construct year variable

We just want a binary variable indicating whether the year is 2017.

```{r}
sample_trips <- sample_trips |>
  mutate(y2017 = ifelse(year == 2017, 1, 0))
```

### Predictor variable summary

Now we summarize the predictor variables.

```{r, message=FALSE}
means <- sample_trips |>
  group_by(year) |>
  summarise(`Income` = mean(income_k)*1000,
            `Vehicles per driver` = mean(veh_per_driver),
            `Presence of non-working father` = mean(non_work_dad),
            `Presence of non-working mother` = mean(non_work_mom),
            Age = mean(age),
            Female = mean(female),
            `Has a younger sibling` = mean(has_lil_sib),
            `Has an older sibling` = mean(has_big_sib),
            `Trip distance (miles)` = mean(TRPMILES),
            `Population density (per sq-mi)` = mean(density)) |>
  t() |>
  as.data.frame() |>
  row_to_names(1) |>
  rownames_to_column("Variable") |>
  rename(`2009 Mean` = `2009`,
         `2017 Mean` = `2017`)

sds <- sample_trips |>
  group_by(year) |>
  summarise(`Income` = sd(income_k)*1000,
            `Vehicles per driver` = sd(veh_per_driver),
            Age = sd(age),
            `Trip distance (miles)` = sd(TRPMILES),
            `Population density (per sq-mi)` = sd(density)) |>
  t() |>
  as.data.frame() |>
  row_to_names(1) |>
  rownames_to_column("Variable") |>
  rename(`2009 Std. Dev.` = `2009`,
         `2017 Std. Dev.` = `2017`)

left_join(means, sds) |>
  select(Variable,
         `2009 Mean`,
         `2009 Std. Dev.`,
         `2017 Mean`,
         `2017 Std. Dev.`) |>
  kable(format.args = list(scientific = FALSE),
        digits = 2)
  
```


## Outcome variable

The outcome variable needs to be numeric for Biogeme. We use the following 
codes:

* 1: Car
* 2: Parental active escorting
* 3: Active without parent

```{r, message=FALSE}
sample_trips <- sample_trips |>
  mutate(with_parent = (ONTD_P1 == "01" & age_01 > 17) |
           (ONTD_P2 == "01" & age_02 > 17) |
           (ONTD_P3 == "01" & age_03 > 17) |
           (ONTD_P4 == "01" & age_04 > 17) |
           (ONTD_P5 == "01" & age_05 > 17) |
           (ONTD_P6 == "01" & age_06 > 17) |
           (ONTD_P7 == "01" & age_07 > 17) |
           (ONTD_P8 == "01" & age_08 > 17) |
           (ONTD_P9 == "01" & age_09 > 17) |
           (ONTD_P10 == "01" & age_10 > 17) |
           (ONTD_P11 == "01" & age_11 > 17) |
           (ONTD_P12 == "01" & age_12 > 17) |
           (ONTD_P13 == "01" & age_13 > 17)) |>
  mutate(mode_ind = case_when(mode == "Car" ~ 1,
                              with_parent & mode == "Walk" ~ 2,
                              with_parent & mode == "Bike" ~ 2,
                              mode == "Walk" ~ 3,
                              mode == "Bike" ~ 3,
                              TRUE ~ -1))

sample_trips |>
  group_by(year, mode_ind) |>
  summarize(n = n()) |>
  pivot_wider(names_from = mode_ind, values_from = n) |>
  rename(Car = `1`,
         `Parental active escorting` = `2`,
         `Active without parent` = `3`) |>
  kable()
```

## Availability variables

Biogeme requires an availability variable. 
We assume that all of the above alternatives are available for all respondents,
so we will also create the following availability variables, and set the values
to 1 (TRUE) for all respondents.

* av_car
* av_par_act
* av_kid_act

```{r}
sample_trips <- sample_trips |>
  mutate(av_car = 1,
         av_par_act = 1,
         av_kid_act = 1)
```

# Select final variable set

```{r}
sample_trips <- sample_trips |>
  ungroup() |>
  select(mode_ind, 
         y2017,
         av_car,
         av_par_act,
         av_kid_act,
         log_inc_k,
         veh_per_driver,
         non_work_dad,
         non_work_mom,
         age,
         female,
         has_lil_sib,
         has_big_sib,
         log_distance,
         log_density)
  
```

# Write csv files

We'll save a file with the full pooled sample and also a separate sample for 
each year.

```{r}
write_csv(sample_trips,
          here("data",
               "model-input",
               "both-years.csv"))

write_csv(sample_trips[sample_trips$y2017==0,],
          here("data",
               "model-input",
               "year2009.csv"))

write_csv(sample_trips[sample_trips$y2017==1,],
          here("data",
               "model-input",
               "year2017.csv"))

```

